Lazy Linearity for a Core Functional Language

### 2025-11-25 popl26.tex
Camera Ready

The implementation of a Linear Core checker does not follow directly from the
type system because Linear Core is not syntax-directed \ROUNDTWO{(the design
aims to be as simple to reason with as possible).}
Specifically, the \ROUNDTWO{features of Linear Core that are not
syntax-directed are essentially the}
splitting of linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
\ROUNDTWO{As standard in substructural typing systems, our rules
  non-deterministically split linear resources as needed.}
\ROUNDTWO{In our implementation,} we thread input/output linear resources through each
derivation~\cite{DBLP:journals/tcs/CervesatoHP00}.
%
To use pattern variables with usage environment $\lctag{\irr{\D}}{K_i}$, the
scrutinee resources must first be $\textsc{Split}$ into fragments. Instead of
guessing which resources need to be split,
the implementation consumes tagged fragments of a resource as needed,
i.e., only when consuming a resource tagged $K_j$ do we $\textsc{Split}$ that
resource on $K$ and consume the $K_j$ fragment.
%
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. We leave the study of the soundness and completeness
of this implementation strategy to future work.
%Further research regarding
%this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}

@@

The results of running the Linear Core GHC plugin on real-world libraries with
linear types are reported in Figure~\ref{fig:core-plugin-res}.
%
%
We compiled the Haskell standard library for programming with linear types,
\texttt{linear-base} (4000 lines, comprising over 100 modules),
\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines),
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines),
\texttt{text-builder-linear}~\cite{cite:text-builder-linear} (1800 lines), and
\texttt{linear-generics}~\cite{cite:linear-generics} (2000 lines), using our
plugin.
%
We count the number of programs accepted by our implementation, where
each top-level binding in a module counts as a program, and every such
binding is checked once per optimisation pass, i.e. we check
all intermediate programs produced by GHC. The total of programs
rejected is counted in the same way, without ever halting compilation when a
program is rejected. We note that our implementation is not performance
conscious and types every intermediate program from scratch.
% , instead of preserving information throughout the pipeline.

>>>>

The implementation of a Linear Core checker does not follow directly from the
type system because Linear Core is not syntax-directed \ROUNDTWO{(the design
aims to be as simple to reason with as possible).}
Specifically, the \ROUNDTWO{features of Linear Core that are not
syntax-directed are essentially the}
splitting of linear resources amongst sub-derivations and
consuming fragments of resources through pattern-bound variables.
\ROUNDTWO{As standard in substructural typing systems, our rules
  non-deterministically split linear resources as needed.}
\ROUNDTWO{In our implementation,} we thread input/output linear resources through each
derivation~\cite{CERVESATO2000133}.
%
To use pattern variables with usage environment $\lctag{\irr{\D}}{K_i}$, the
scrutinee resources must first be $\textsc{Split}$ into fragments. Instead of
guessing which resources need to be split,
the implementation consumes tagged fragments of a resource as needed,
i.e., only when consuming a resource tagged $K_j$ do we $\textsc{Split}$ that
resource on $K$ and consume the $K_j$ fragment.
%
Finally, our implementation must infer the usage environments of binders in
a recursive let group before using them to typecheck the let body. We use a
naive $O(n^2)$ algorithm (where $n$ is the number of let bindings) to determine
these usage environments. We leave the study of the soundness and completeness
of this implementation strategy to future work.
%Further research regarding
%this inference challenge is discussed in Section~\ref{sec:future-work}.
%
%
%
% We use the following techniques to tackle the \emph{non-syntax-directedness} of
% Linear Core:
%
% \begin{itemize}

@@

The results of running the Linear Core GHC plugin on real-world libraries with
linear types are reported in Figure~\ref{fig:core-plugin-res}.
%
%
We compiled the Haskell standard library for programming with linear types,
\texttt{linear-base} (4000 lines, comprising over 100 modules),
\texttt{linear-smc}~\cite{10.1145/3471874.3472980} (1500 lines),
\texttt{priority-sesh}~\cite{10.1145/3471874.3472979} (1400 lines),
\texttt{text-builder-linear} \cite{cite:text-builder-linear} (1800 lines), and
\texttt{linear-generics}~\cite{cite:linear-generics} (2000 lines), using our
plugin.
%
We count the number of programs accepted by our implementation, where
each top-level binding in a module counts as a program, and every such
binding is checked once per optimisation pass, i.e.~we check
all intermediate programs produced by GHC. The total of rejected programs
is counted in the same way, without halting compilation. We note that our implementation is not performance
conscious and types every intermediate program from scratch.
% , instead of preserving information throughout the pipeline.
